---
title: "Gggplot2 Internals - An Unofficial Reference"
author: "Brodie Gaslam"
output:
    rmarkdown::html_vignette:
        toc: true
        css:
          - !expr file.path(system.file(package="diffobj"),"css","diffobj.css")
          - styles.css
vignette: >
  %\VignetteIndexEntry{extensions}
  %\VignetteEngine{knitr::rmarkdown}
  %\usepackage[utf8]{inputenc}
---

```{r global_options, echo=FALSE}
options(
  digits=2, diffobj.pager="off", crayon.enabled=TRUE,
  # diffobj.format="ansi256", diffobj.brightness="light",
  diffobj.format="html", diffobj.disp.width=120,
  diffobj.ignore.white.space=TRUE
)
sgr_wrap <- function(x, options){
  paste0(
    "<pre class=\"r-output\"><code>",
    fansi::sgr_to_html(x = htmltools::htmlEscape(x)),
    "</code></pre>"
  )
}
# knitr::knit_hooks$set(output=sgr_wrap)
knitr::opts_chunk$set(error=TRUE, comment=NA)
library(ggbg)
```

## Introduction

This document is an unofficial reference for developers looking to extend
`ggplot2`.  It supplements existing documentation by providing more
detail about how the extension mechanism interacts with other parts of the
`ggplot2` plot building process.  It is based on examination of the sources of
`ggplot2` circa 5/2018 (version 3.0 pre-release), and may become out of date as
`ggplot2` evolves.  Please be aware that by virtue of this documentation being
unofficial, behavior it documents that is not also documented officially may be
changed by the `ggplot2` maintainers without warning.

We assume the reader is familiar with using `ggplot2` for generating plots, and
that the reader has previously reviewed:

* `?ggproto` for an introduction to the `ggplot2` object oriented programming
  system.
* `?ggplot2::ggplot2-ggproto` for the major `ggproto` objects used by `ggplot2`.
* `vignette(package='ggplot2', 'extending-ggplot2')` for the official
  documentation on extending `ggplot2`.

There are four major sections to this document:

1. [Overview](#overview): a look at the process of converting data
   and user plot specifications into a plot.  This section provides
   context for the reference sections that follow and is the only section
   intended to be read from beginning to end.
2. [Method Outline](#method-outline): listing of all the key function calls
   used to generate the plot, which gives you a sense for where
   your extension fits in relation to the rest of the process.
3. [Method Detail](#method-details): in-depth look at the key [`ggproto`
   methods](#ggproto-method) used to generate the plot.
4. [Definitions](#definitions): definitions of key terms.

## Overview

### It's All in the Printing

Four major steps are required to produce a `ggplot`:

1. Specification: user specifies what the plot should look like with
   `ggplot(...) + geom_*(...)`, etc..
2. Build: manipulate the data as directed by the user plot specification (apply
   stats, position adjustments, scale transformations, map to
   [aesthetics](#aesthetic) etc.).
3. Render: translate build plot data into [`grid`](#grid) [graphical objects
   (grobs)](#grob).
4. Display: output the plot to screen.

Step 1 is done directly by the user and will be familiar to all `ggplot2` users.
Steps 2-4 are usually triggered by the printing of a `ggplot2` object, as shown
by this simplified version of `print.ggplot`:

```{r eval=FALSE}
print.ggplot <- function (x, ...) {
  data <- ggplot_build(x)        # step 2: build
  gtable <- ggplot_gtable(data)  # step 3: create grobs
  grid.draw(gtable)              # step 4: display grobs
}
```
Almost all the work `ggplot2` does is in these last three steps.  It is helpful
to understand what happens in them when extending `ggplot2`.  This document
focuses on the [`ggplot_build`](#ggplot_build) and
[`ggplot_gtable`](#ggplot_gtable) steps.

### Extension Mechanism

Before we get into the nitty-gritty of `ggplot_build` and `ggplot_gtable`, we
will review the `ggplot2` extension mechanism.  `ggplot2` is extended by
defining [`ggproto` objects](#ggproto) that inherit from the existing
[`Geom`](#ggproto-geom), [`Stat`](#ggproto-stat),
[`Position`](#ggproto-position), [`Scale`](#ggproto-scale),
[`Coord`](#ggproto-coord), [`Facet`](#ggproto-facet), and other [`ggproto`
objects](#ggproto).  [`ggproto` objects](#ggproto) contain both data and
functions (a.k.a. [methods](#ggproto-method)).  This allows us to override
existing functionality by providing `ggproto` objects with custom
[methods](#ggproto-method).

If we want to define a new `Geom` "x", we create a `GeomX` `ggproto` object that
extends `Geom` and modifies the relevant methods.  We then provide a
corresponding [layer function](#layer-function) that associates it with a
[`Layer` `ggproto` object](#ggproto-layer) that can be added to the `ggplot`.

For examples on how to do this see `vignette(package='ggplot2',
'extending-ggplot2')`.

### Plot Structure

A `ggplot` is a nested structure of [`ggproto`](#ggproto) and other objects.
It is initialized by the `ggplot(...) + ...` specification step.  This is a
simplified version of the structure:

```
plot (list)
+ Coord (ggproto)
+ Facet (ggproto)
+ Scales (ggproto) can contain multiple Scale objects
    + Scale (ggproto)
+ layers (list) can contain multiple Layer objects
    + Layer (ggproto)
      + data (data.frame)
      + Geom (ggproto)
      + Stat (ggproto)
      + Position (ggproto)
      + ...
+ ...
```

This implies a plot:

* Can have multiple [layers](#layer).
* Can have multiple [scales](#ggproto-scale), however the set of scales is shared
  across every [layer](#layer) since layers themselves do not have scales.
* Has one coordinate system.
* Has one set of facets.
* Has independent data, [`Geom`](#ggproto-geom), [`Stat`](#ggproto-stat), and
  [`Position`](#ggproto-position) for each layer.

Additionally this implies every layer maps onto the same space since layers rely
on the plot-level scales and facets.

The portion of the structure you are most likely to interact with when extending
`ggplot2` are the sub-objects of the [`Layer`](#ggproto-layer) object.  Within
each [`Layer` object](#ggproto-layer) there is an additional level of implicit
structure we did not show above. The [`ggplot_build`](#ggplot_build) process
adds [`PANEL`](#panel) and `group` variables to the [`Layer$data`](#layer-data)
`data.frame`.  This creates groups of rows that each correspond to a particular
[panel](#panel) or group within that [panel](#panel).  The pseudo structure of a
[`Layer`](#ggproto-layer) is then:

```
Layer
    + data
    + Panel
        + Group
    + Geom
    + Stat
    + Position
    + ...
```

We say "pseudo" structure because there are no actual distinct objects that
represent panels and groups.  They are implied by the `group` and `PANEL`
variables of the [layer data](#layer-data).

As a result of this structure many steps in the [`ggplot_build`](#ggplot_build)
process follow this pseudo code pattern:

```
For Each Layer:
    For Each Panel:
        For Each Group:
            Do Something (usually to the data).
```

Layer-related [`ggproto` objects](#ggproto) reflect this in their own structure
by the presence of `$*_layer`, `$*_panel`, and `$*_group`
[methods](#ggproto-method).

### Example Plot

Throughout the rest of this document we will use the following plot for
illustrative purposes:

```{r, eval=FALSE}
dat <- data.frame(
  loc=1:4,
  val=5:8,
  grp=c('A', 'B', 'B', 'C'),
  pnl=c('P_ONE', 'P_ONE', 'P_TWO', 'P_TWO')
)
library(ggplot2)
p <- ggplot(dat, aes(x=loc)) +
  stat_bin(bins=1, aes(weight=val), fill="lightgray") +
  geom_point(aes(y=val, fill=grp), position="jitter", shape=21, size=4) +
  facet_wrap(~pnl, scales="free_x") +
  scale_x_reverse()
```
```{r, eval=FALSE, echo=FALSE}
p <- ggplot(dat, aes(x=loc)) +
  stat_bin(bins=1, aes(weight=val), fill="lightgray") +
  geom_point(
    aes(y=val, fill=grp), position=position_jitter(seed=24), shape=21, size=4
  ) +
  facet_wrap(~pnl, scales="free_x") +
  scale_x_reverse()
```

## Method Outline

### ggplot_build

`ggplot_build` starts with the input data and the [plot
specification](#plot-spec).  It then proceeds through several steps that
sequentially transform the data by adding and/or recomputing variables, or in
the case of statistics recomputing the entire data.  In addition to transforming
the data, the build process also updates and computes meta data for scales,
facets, etc.  The work is carried out by the methods of the [`ggproto`
objects](#ggproto) provided implicitly or explicitly via the [plot
specification](#plot-spec).

An outline of the methods called by `ggplot_build` follows.  This is
purposefully simplified by omitting some of the calls, or the control structures
around some of the calls.  For example, some methods are called on every
layer and we do not explicitly show that here, though we do so in the [details
section](#method-details).

The outline is intended to give a rough idea of what happens in what order.  It
also acts as an index to [more comprehensive documentation](#method-details) for
the key methods.  You are encouraged to use this section as a reference; search
for the [`ggproto` objects](#ggproto) you are extending.

* [`Layer$layer_data`](#layerlayer_data): assign data to each layer.
* `Layout$setup`:
    * [`Facet$setup_params`](#setup_params-setup_data): unused by common `Facet`
      objects.
    * [`Facet$setup_data`](#setup_params-setup_data): unused by common `Facet`
      objects.
    * [`Coord$setup_params`](#setup_params-setup_data): unused by common `Coord`
      objects.
    * [`Coord$setup_data`](#setup_params-setup_data): unused by common `Coord`
      objects.
    * [`Facet$compute_layout`](#facetcompute_layout): generate facet meta data.
    * `Coord$setup_layout`: unused by common `Coord` objects, provides an
      opportunity to adjust the [facet layout](#facetcompute_layout).
    * [`Facet$map_data`](#facetmap_data): add the `PANEL` variable that maps
      data rows to facets.
* [`Layer$compute_aesthetics`](#layercompute_aesthetics): compute
  [aesthetics](#aesthetic), add `group` variable.
* [`Scale$transform_df`](#scaletransform_df): apply scale transformations to
  [aesthetics](#aesthetic).
* `Layout$train_position`: initialize and [train](#train-scale) [position
  scales](#ggproto-scale-position).
* [`Layout$map_position`](#layoutmap_position): [map](#map-scale) [position
  scales](#ggproto-scale-position).
* `Layer$compute_statistic`: recompute data with statistics.
    * [`Stat$setup_params`](#setup_params-setup_data): adjust parameters.
    * [`Stat$setup_data`](#setup_params-setup_data): adjust data.
    * [`Stat$compute_layer`](#stat-compute): apply statistic to each layer.
    * [`Stat$compute_panel`](#stat-compute): apply statistic to each panel.
    * [`Stat$compute_group`](#stat-compute): apply statistic to each group.
* [`Layer$map_statistic`](#layermap_statistic): re-computes and rescales
  aesthetics using the data produced by [`Stat$compute_layer`](#stat-compute).
* `scales_add_missing`: adds default [position scales](#ggproto-scale-position)
  to each layer if they are missing.
* `Layer$compute_geom_1`: prepare `Geom` data for each layer.
    * [`Geom$setup_data`](#setup_params-setup_data): adjust data.
    * `check_required_aesthetics`: check that the required aesthetics are
      present in the [layer data](#layer-data) or as [parameter
      aesthetics](#param-aesthetic).
* `Layer$compute_position`: apply position adjustments.
    * [`Position$setup_params`](#setup_params-setup_data): adjust parameters.
    * [`Position$setup_data`](#setup_params-setup_data): adjust data.
    * [`Position$compute_layer`](#position-compute): adjust positions for
      layer.
    * [`Position$compute_panel`](#position-compute): adjust positions for
      panel.
* `Layout$reset_scales`: remove any previous [scale training](#train-scale).
* `Layout$train_position`: [re-train](#train-map-scale) the [position
  scales](#ggproto-scale-position).
* [`Layout$setup_panel_params`](#layoutsetup_panel_params): compute meta-data
  for panels.
* [`Layout$map_position`](#layoutmap_position): re-map [position
  scales](#ggproto-scale-position).
* [`scales_train_df`](#train-map-scale): [train](#train-scale)
  non-[position scales](#ggproto-scale-position).
* [`scales_map_df`](#train-map-scale): [map](#map-scale)
  non-[position scales](#ggproto-scale-position).
* `Layer$compute_geom_2`: add defaults and aesthetic parameters.
    * [`Geom$use_defaults`](#geomuse_defaults).
* `Layer$finish_statistics`:
    * `Stat$finish_layer`: typically unused, gives stat one more chance to
      modify data after it is [mapped](#map-scale).
* `Layout$finish_data`:
    * `Facet$finish_data`: typically unused, gives `Facet` one last chance to
      modify data with access to the [layout](#layoutlayout) meta data and the
      [position scales](#ggroto-scales-position).

### ggplot_gtable

In this step we [render](#render) the `ggplot_built` object produced by
[`ggplot_build`](#ggplot_build) into the [`grobs`](#grob) that will ultimately
be turned into shapes, colors, and text on screen.  Unlike
[`ggplot_build`](#ggplot_build), only a small portion of the work is done by
[`ggproto` methods](#ggproto-method) that can be extended by users, so there is
not as much for us to document here.

* `Layer$draw_geom`: render the layer geom [grobs](#grob) and arrange them in a
  [`gtable`](#gtable)
    * [`Coord$transform`](#coordtransform): One of the
      [`Geom$draw_*`](#geom-draw) methods should apply coordinate
      transformations.
    * [`Geom$draw_layer`](#geom-draw): render [grobs](#grob) for layer.
    * [`Geom$draw_panel`](#geom-draw): render [grobs](#grob) for panel.
    * [`Geom$draw_group`](#geom-draw): render [grobs](#grob) for group
* `Layout$render`: render the rest of the [panel](#panel) and [facet](#facet)
  elements, combine them with the layer geom [grobs](#grob), and return as a
  [`gtable`](#gtable).
* Generate legend and add it to the layout [`gtable`](#gtable).
* Add tag, title, subtitle, and margins to the [`gtable`](#gtable).

## Method Details

### Layer$layer_data

For each layer, this method assigns the data to be used, which is either the
data specified in the `geom_*` / `stat_*` [layer function](#layer-function) call
if any, or if not the data specified in the `ggplot()` call.  Because neither
layer in the [example plot](#example-plot) had data they both acquire the [plot
data](#plot-data):

```
 $ :'data.frame':	4 obs. of  4 variables:          # STAT LAYER
  ..$ loc: int [1:4] 1 2 3 4
  ..$ val: int [1:4] 5 6 7 8
  ..$ grp: Factor w/ 3 levels "A","B","C": 1 2 2 3
  ..$ pnl: Factor w/ 2 levels "P_ONE","P_TWO": 1 1 2 2
 $ :'data.frame':	4 obs. of  4 variables:          # GEOM LAYER
  ..$ loc: int [1:4] 1 2 3 4
  ..$ val: int [1:4] 5 6 7 8
  ..$ grp: Factor w/ 3 levels "A","B","C": 1 2 2 3
  ..$ pnl: Factor w/ 2 levels "P_ONE","P_TWO": 1 1 2 2
```

The [layer data](#layer-data) is then kept in a list that is updated throughout
the [`ggplot_build`](#ggplot_build) process.

### Object\$setup_params And Object\$setup_data {#setup_params-setup_data}

#### Description

`ggplot2` has a complex [plot specification](#plot-spec) system.  In
most cases, the users typed commands are relatively simple, but `ggplot2` then
needs to set or compute default values for derived variables and parameters.  To
assist with this most of the [`ggproto` objects](#ggproto) have `$setup_params`
and `$setup_data` methods.  These are often called after some of the initial
setup has been done (e.g. [assigning layer data](#layerlayer_data)) so that they
can use this information to update the data or parameters.

For example, `Facet$setup_params` is called with a list containing the plot data
and the data for each layer after.  This gives the method the opportunity to
update the parameters in the context of the entirety of the data.

The `$setup_data` methods provide an opportunity to update the data, also after
the data has been assigned to layers, and after the parameters have been updated
with the `$setup_params` methods.  An example is how some `Geoms` will add
`xmin` and `xmax` [aesthetics](#aesthetic) computed from existing `x` and
`width` aesthetics with the `$setup_data` method.

#### Signature

* `$setup_params([self], data, params)`
* `$setup_data([self], data, params)`

The `self` parameter, if present, can appear anywhere in the signature.  The
other parameters must be in the relative order shown above.  See [`ggproto`
methods](ggproto-method).

#### Parameters

* `self`: optional parameter, if present will point to the `ggproto` object the
  method belongs to.
* `data`: the layer data.  Depending on which `ggproto` object the `$setup_*`
  method belongs to it may have undergone scale transformation and/or mapping.
  See the [method outline](#method-outline) to get an idea of the state of the
  data by the time it gets to the `$setup_*` method in question.
* `params`: a list of the parameters provided to the [layer
  function](#layer-function) that match the parameters of the
  `$[compute|draw]_panel` of `self`, and if that method has a `...` parameter,
  then also the parameters of the `$[compute|draw]_group` method.

#### Data Manipulation

For `$setup_params`, a list of the parameters associated with the corresponding
[`ggproto`](#ggproto) object.

For `$setup_data`, the updated data.

#### See Also

Some of the default methods are documented in `?'ggplot2-ggproto'`.  The
definition of [ggproto methods](#ggproto-method) for details about the `self`
parameter, [layer functions](#layer-function) for details about how the `params`
list is generated, the [stat](#stat-compute) and
[position](#position-compute) computation methods, or the [geom
draw](#geom-draw) methods for methods that operate on the data after it has been
setup.

### Facet$compute_layout

Assigns a panel, row, column, and [position scale](#ggproto-scale-position) ids
to each facet.  It is possible to have multiple [position
scales](#ggproto-scale-position) in a plot when the scales are "free".

This data is stored in the `$layout` <a name='layout-layout'></a> member of the
[`Layout` `ggproto`](#ggproto-layout) object.  This is what it looks like for
our [example plot](#example-plot):

```
  PANEL ROW COL   pnl SCALE_X SCALE_Y
1     1   1   1 P_ONE       1       1
2     2   1   2 P_TWO       2       1
```

Our data has two facets (panels) arranged across two columns and in one row.
There is one `y` scale, and two `x` scales because with set "free_x" in the
facet specification for the plot.

### Facet$map_data

For each layer, add the [`PANEL`](#panel) variable to the data.  Here we show
the effect it has on the data of the first layer of our [example
plot](#example-plot), although the effect will be the same on the second as they
both share the [plot data](#plot-data):

```{r, echo=FALSE, results="asis"}
facet.map.in <- structure(list(loc = 1:4, val = 5:8, grp = structure(c(1L, 2L,
2L, 3L), .Label = c("A", "B", "C"), class = "factor"), pnl = structure(c(1L,
1L, 2L, 2L), .Label = c("P_ONE", "P_TWO"), class = "factor")), class = "data.frame", row.names = c(NA,
-4L))
facet.map.out <- structure(list(loc = 1:4, val = 5:8, grp = structure(c(1L, 2L,
2L, 3L), .Label = c("A", "B", "C"), class = "factor"), pnl = structure(c(1L,
1L, 2L, 2L), .Label = c("P_ONE", "P_TWO"), class = "factor"),
    PANEL = structure(c(1L, 1L, 2L, 2L), .Label = c("1", "2"), class = "factor")), row.names = c(NA,
4L), class = "data.frame")

diffobj::diffPrint(facet.map.in, facet.map.out)
```

### Layer$compute_aesthetics

For each layer, evaluates all the [mapped aesthetic](#mapped-aesthetic)
`aes(...)` expressions and replaces the data variables with the resulting
aesthetics.  [Stat aesthetics](#stat-aesthetic) are evaluated later as part of
the [stat computation](#stat-compute) process.

This step also infers the `group` aesthetic from the interaction of categorical
[mapped aesthetics](#mapped-aesthetic).

Here we show the effect on the `geom_point` [layer data](#layer-data):

```{r, echo=FALSE, results="asis"}
comp.aes.in <- facet.map.out
comp.aes.out <- structure(list(y = 5:8, fill = structure(c(1L, 2L, 2L, 3L), .Label = c("A",
"B", "C"), class = "factor"), x = 1:4, PANEL = structure(c(1L,
1L, 2L, 2L), .Label = c("1", "2"), class = "factor"), group = structure(c(1L,
2L, 2L, 3L), n = 3L)), row.names = c(NA, -4L), class = "data.frame")
diffobj::diffPrint(comp.aes.in, comp.aes.out)
```

The variables have changed to be the [aesthetics](#aesthetic), and we gain the
`group` variable.

### Scale$transform_df

This is applies scale transformations to each layer.  In our example, we use
`scale_x_reverse`.  Here is the effect on the data for the `geom_point` layer:

```{r, echo=FALSE, results="asis"}
trans.in <- comp.aes.out
trans.out <- structure(list(x = -1:-4, y = 5:8, fill = structure(c(1L, 2L,
2L, 3L), .Label = c("A", "B", "C"), class = "factor"), PANEL = structure(c(1L,
1L, 2L, 2L), .Label = c("1", "2"), class = "factor"), group = structure(c(1L,
2L, 2L, 3L), n = 3L)), class = "data.frame", row.names = c(NA,
-4L))

diffobj::diffPrint(trans.in, trans.out)
```

The transformation is actually applied by  `ggplot2:::scales_transform_df` which
calls `Scale$transform_df` with each of the scales of the plot.
`Scale$transform_df` then `Scale$transform` on all the aesthetics that are
associated with the scale.  You can check which aesthetics are associated with a
scale with the `$aesthetics` member of the scale:

```{r}
scale <- ggplot2::scale_x_continuous()
scale$aesthetics
```

In this case the only `aesthetic` transformed is `x`, but if `xmin` and `xmax`
had been present they too would have been transformed.

For our numeric `x` aesthetic the transformation was achieved by making the
numbers negative, effectively reversing them.  When axes labels and such are
displayed, the transformation will be applied in the inverse direction on them.

### Layout$map_position

For each layer, [map](#map-scale) data with [position scales](#position-scale).

For continuous variables this in theory would rescale the continuous position
aesthetics (e.g. `x`, `xmin`, `xmax`, `y`, etc.) into the display range (i.e.
0-1), but because this is handled at the [coordinate
transformation](#coordtransform) step for [position
aesthetics](#position-aesthetic) the only thing does is handle NAs and convert
integer aesthetics to numeric.

Discrete variables are mapped onto integers for each value.  This only happens
once even though the build process calls `Layout$map_position` twice.  The
second time it will only be applied to aesthetics that have not already been
converted (i.e. those added by a [`Stat`](#ggproto-stat)).

### Stat$compute_layer/panel/group {#stat-compute}

#### Description

These methods are responsible for applying a statistic to the [layer
data](#layer-data) and returning the result as the new layer data.  Typically
you would modify one of the layer, panel, or group methods, listed here in order
of increasing specificity.  If possible, you should modify the most specific
method (e.g. the group method) as that requires the least work.  If you need
more control over the process you can override one of the more general methods.
If you do this, for example by overriding the panel method, you will be
responsible for splitting the data into groups, processing the groups, and
re-assembling the data.

`$compute_layer` is called on each layer of the plot.  The default
`$compute_layer` method will call `$compute_panel` on each [panel](#panel).
Finally the default `$compute_panel` method will call `$compute_group` on each
group in the [panel](#panel).

#### Signature

* `$compute_layer([self], data, params, layout)`
* `$compute_panel([self], data, scales, ...)`
* `$compute_group([self], data, scales)`

The `self` parameter, if present, can appear anywhere in the signature.  The
other parameters must be in the relative order shown above.  If you override
`$compute_panel` and `$compute_group` you may add additional parameters after
the `scales` parameter.  These parameters will be automatically matched by and
passed on from the [layer function](#layer-function).  See [`ggproto`
methods](#ggproto-method).

#### Parameters

* `self`: optional parameter, if present will point to the [`Stat`
  `ggproto`](#ggproto-stat) object the method belongs to.
* `data`: the [layer data](#layer-data) for `$compute_layer`, a subset of the
  layer data corresponding to a [panel](#panel) for `$compute_panel`, or a
  subset of the panel data corresponding to a group for `$compute_group`.
* `params`: the parameter list as matched to the `$compute_panel/group` methods
  by the [layer function](#layer-function).
* `layout`: the [`Layout` `ggproto`](#ggproto-layout) object for the plot,
  provided mostly so that `$compute_layer` can extract the [position
  scales](#ggproto-scale-position) to
  pass on to `$compute_panel` and `$compute_group`.
* `scales`: a list with the [position scales](#ggproto-scale-position) for the
  panel.  This can be useful if values in `params` need to be transformed by the
  scales (e.g. as with breaks corresponding to `x` values that are passed in
  `params`).  `scales[['x']]` contains the `x` scale, and `scales[['y']]`
  contains the `y` scale.

#### Data Manipulation

Statistics can completely change the data.  Here is what the value of the `data`
parameter for `Stat$compute_layer` looks like in [example plot](#example-plot)
for the `stat_bin` layer:

```{r, echo=FALSE}
stat.comp.in <- structure(list(x = c(-1, -2, -3, -4), weight = 5:8, PANEL = structure(c(1L,
1L, 2L, 2L), .Label = c("1", "2"), class = "factor"), group = c(-1L,
-1L, -1L, -1L)), row.names = c(NA, -4L), class = "data.frame")
stat.comp.in
```

And if you have not modified the `$compute_layer` method, then the value of the
`data` parameter for `$compute_panel` looks like (for the first panel):

```
   x weight PANEL group
1 -1      5     1    -1
2 -2      6     1    -1
```

Finally, if you did not modify the `$compute_panel` method (as is the case with
`StatBin`), the input into `$compute_group` will be the `$compute_panel` data
further subset by group.  In this case this is the same as the panel data as
there is only one group.

In the [example plot](#example-plot) `StatBin` overrides the `$compute_group`
method.  The result for the first (and only) group of the first panel is then:

```
  count    x xmin xmax width density ncount ndensity
1    11 -1.5   -2   -1     1       1      1        1
```

`$compute_group` changed number of rows, and both added and dropped
[`aesthetic`](#aesthetic) variables.

If you only override the `$compute_group` method, `$compute_panel` will add back
any variables that are lost by `$compute_group`, although it does so by
recycling the first value of those variables.  In other words, dropped
aesthetics will take on the value of the first row of each group in the pre-stat
application [layer data](#layer-data).  Since groups are often generated
implicitly from the interaction of all the categorical [aesthetics](#aesthetic),
and these are the aesthetics that typically get dropped, this generally produces
the desired outcome.

In this case `$compute_panel` adds back the `fill`, `PANEL`, and `group`
variables that were dropped by `$compute_group`.  If you override
`$compute_panel` then you will be responsible for doing this.

The final output after `$compute_panel` and `$compute_layer` stitch back the
groups and panels together:

```{r echo=FALSE}
stat.comp.out <- structure(list(count = c(11, 15), x = c(-1.5, -3.5), xmin = c(-2,
-4), xmax = c(-1, -3), width = c(1, 1), density = c(1, 1), ncount = c(1,
1), ndensity = c(1, 1), PANEL = structure(1:2, .Label = c("1",
"2"), class = "factor"), group = c(-1L, -1L)), class = "data.frame", row.names = c(NA,
-2L))
```

If we compare to the input to `$compute_layer` we can see that several variables
(the [stat aesthetics](#stat-aesthetic)) were added to the data:

```{r echo=FALSE, results="asis"}
diffobj::diffStr(stat.comp.in, stat.comp.out)
```

Ultimately, this step should produce data that once processed by
[`Layer$map_statistic`](#layermap_statistic) has all the required aesthetics
required by the [`Geom`](#ggproto-geom) associated with the layer.

### Layer$map_statistic

#### Description

Since [`Stat$compute_layer`](#stat-compute) creates new data, we need to
re-apply some of the early computations and transformations on the aesthetics.
`$map_statistic` effectively applies the following to our stat-recomputed data:

* [`Layer$compute_aesthetics`](#layercompute_aesthetics) (kind of).
* [`Scale$transform_df`](#scaletransform_df).

`$map_statistic` does not actually call `$compute_aesthetic`, but it does most
of what `$compute_aesthetic` does.  There are differences, for example
in this method [stat aesthetics](#stat-aesthetic) are evaluated, whereas
they are not by `$compute_aesthetic`.

#### Signature

* `$map_statistic([self], data, plot)`

The `self` parameter, if present, can appear anywhere in the signature.  The
other parameters must be in the relative order shown above.  See [`ggproto`
methods](#ggproto-method).

#### Parameters

* `self`: optional parameter, if present will point to the `Stat` `ggproto`
  object the method belongs to.
* `data`: the data for a layer after the statistic has been applied.
* `plot`: the `Plot` `ggproto` object.

#### Data Manipulation

The data, with [stat](#stat-aesthetic) and other aesthetics specified by
the [`Stat`](#ggproto-stat) [layer function](#layer-function) that were dropped
by the statistic added back.

```{r, echo=FALSE, results="asis"}
stat.map.in <- stat.comp.out
stat.map.out <- structure(list(y = c(11, 15), count = c(11, 15), x = c(-1.5,
-3.5), xmin = c(-2, -4), xmax = c(-1, -3), width = c(1, 1), density = c(1,
1), ncount = c(1, 1), ndensity = c(1, 1), PANEL = structure(1:2, .Label = c("1",
"2"), class = "factor"), group = c(-1L, -1L)), class = "data.frame", row.names = c(NA,
-2L))

diffobj::diffStr(stat.map.in, stat.map.out)
```
### Position$compute_layer/panel {#position-compute}

#### Description

These methods adjust [position aesthetics](#position-aesthetic) in the data for
groups of elements that are considered to overlap.  For example,
both `position_dodge` and `position_stack` group data by `xmin` aesthetic value
(either user provided, or derived from `x` and `width` or some such), and then
resolve any `xmin` overlaps by allocating the available width for each distinct
`group` within, or modifying `ymin` / `ymax` values so elements will stack.

`$compute_layer` is called on each layer of the plot, and the default method
then calls `$compute_panel` on each panel in the layer.  Unlike with the [stat
computation](#stat-compute) methods there is not a default `$compute_group`
method since typically the position adjustment must be made accounting for all
the groups positions simultaneously.

#### Signature

* `$compute_layer([self], data, params, layout)`
* `$compute_panel([self], data, params, scale)`

The `self` parameter, if present, can appear anywhere in the signature.  The
other parameters must be in the relative order shown above.  See [`ggproto`
methods](#ggproto-method).

#### Parameters

* `self`: optional parameter, if present will point to the `Stat` `ggproto`
  object the method belongs to.
* `data`: the [layer data](#layer-data) for `$compute_layer`, a subset of the
  layer data corresponding to a [panel](#panel) for `$compute_panel` .
* `params`: the parameter list as matched to the `$compute_panel/group` methods
  by the [layer function](#layer-function).
* `layout`: the [`Layout` `ggproto`](#ggproto-layout) object for the plot,
  provided mostly so that `$compute_layer` can extract the [position
  scales](#ggproto-scale-position) to
  pass on to `$compute_panel` and `$compute_group`.
* `scales`: a list with the [position scales](#ggproto-scale-position) for the
  panel.  This can be useful if values in `params` need to be transformed by the
  scales (e.g. as with breaks corresponding to `x` values that are passed in
  `params`).  `scales[['x']]` contains the `x` scale, and `scales[['y']]`
  contains the `y` scale.

#### Data Manipulation

The data with positions adjusted.  Here is how `$compute_layer` changes the
layer data for the `geom_point` layer from our [example plot](#example-plot)
that uses `position_jitter`:

```{r, echo=FALSE, results="asis"}
pos.comp.in <- structure(list(x = c(-1, -2, -3, -4), y = c(5, 6, 7, 8), fill = structure(c(1L,
2L, 2L, 3L), .Label = c("A", "B", "C"), class = "factor"), PANEL = structure(c(1L,
1L, 2L, 2L), .Label = c("1", "2"), class = "factor"), group = structure(c(1L,
2L, 2L, 3L), n = 3L)), row.names = c(NA, -4L), class = "data.frame")
pos.comp.out <- structure(list(x = c(-1.16594083365053, -2.22008708454669, -2.83662163671106,
-3.9848822819069), y = c(5.13009568769485, 6.33635502737016,
6.8237884586677, 8.21105636898428), fill = structure(c(1L, 2L,
2L, 3L), .Label = c("A", "B", "C"), class = "factor"), PANEL = structure(c(1L,
1L, 2L, 2L), .Label = c("1", "2"), class = "factor"), group = structure(c(1L,
2L, 2L, 3L), n = 3L)), row.names = c(NA, -4L), class = "data.frame")
diffobj::diffPrint(pos.comp.in, pos.comp.out)
```

This adjust the `x` and `y` aesthetics, though it would also adjust any other
[position aesthetics](#position-aesthetic) if they existed in the data.

If you overrode the `$compute_panel` method instead of `$compute_layer` as
`position_jitter` does, the inputs would be the same, except subset by the
`PANEL` variable.

### Layout$setup_panel_params

Generates a list that, for each [panel](#panel), contains a list of the key meta
data for the [position scales](#ggproto-scale-position), including the range,
labels, location of major and minor breaks, and axis order, amongst others.  The
list is then stored in `Layout$panel_params`, by [reference](#reference).  Here
is what this list looks like in our [example plot](#example-plot):

```
List of 2
 $ :List of 14
  ..$ x.range       : num [1:2] -2.281 -0.939
  ..$ x.labels      : chr [1:6] "1.00" "1.25" "1.50" "1.75" ...
  ..$ x.major       : num [1:6] 0.955 0.768 0.582 0.396 0.209 ...
  ..$ x.minor       : num [1:11] 0.955 0.861 0.768 0.675 0.582 ...
  ..$ x.major_source: num [1:6] -1 -1.25 -1.5 -1.75 -2 -2.25
  ..$ x.minor_source: num [1:11] -1 -1.12 -1.25 -1.38 -1.5 ...
  ..$ x.arrange     : chr [1:2] "secondary" "primary"
  ..$ y.range       : num [1:2] -0.75 15.75
  ..$ y.labels      : chr [1:4] "0" "5" "10" "15"
  ..$ y.major       : num [1:4] 0.0455 0.3485 0.6515 0.9545
  ..$ y.minor       : num [1:7] 0.0455 0.197 0.3485 0.5 0.6515 ...
  ..$ y.major_source: num [1:4] 0 5 10 15
  ..$ y.minor_source: num [1:7] 0 2.5 5 7.5 10 12.5 15
  ..$ y.arrange     : chr [1:2] "primary" "secondary"
 $ :List of 14
  ..$ x.range       : num [1:2] -4.06 -2.78
  ..$ x.labels      : chr [1:5] "3.00" "3.25" "3.50" "3.75" ...
  ..$ x.major       : num [1:5] 0.8269 0.6315 0.4362 0.2408 0.0455
  ..$ x.minor       : num [1:10] 0.925 0.827 0.729 0.632 0.534 ...
  ..$ x.major_source: num [1:5] -3 -3.25 -3.5 -3.75 -4
  ..$ x.minor_source: num [1:10] -2.88 -3 -3.12 -3.25 -3.38 ...
  ..$ x.arrange     : chr [1:2] "secondary" "primary"
  ..$ y.range       : num [1:2] -0.75 15.75
  ..$ y.labels      : chr [1:4] "0" "5" "10" "15"
  ..$ y.major       : num [1:4] 0.0455 0.3485 0.6515 0.9545
  ..$ y.minor       : num [1:7] 0.0455 0.197 0.3485 0.5 0.6515 ...
  ..$ y.major_source: num [1:4] 0 5 10 15
  ..$ y.minor_source: num [1:7] 0 2.5 5 7.5 10 12.5 15
  ..$ y.arrange     : chr [1:2] "primary" "secondary"
```

### Train/Map Scales {#train-map-scale}

#### Description

`scales_train_df` runs the [scale training](#train-scale) on every aesthetic
corresponding to the provided scales (non-[position
scales](#ggproto-scale-position) in `ggplot_build`) across every layer.

`scales_map_df` [maps](#map-scale) every aesthetic corresponding to the provided
scales (non-[position scales](#ggproto-scale-position) in `ggplot_build`) across
every layer.  `scales_map_df` calls internally calls `Scale$map_df` which itself
calls `Scale$map`.

#### Data Manipulation

`scales_train_df` only updates the scale meta data.  `scales_map_df` will modify
the data.  In our [example plot](#example-plot) only the `geom_point` layer has
[mapped](#mapped-aesthetic) non-[position aesthetics](#position-aesthetic)
(`fill`) so it is the only one modified by `scales_map_df`:

```{r, echo=FALSE, results="asis"}

map.scale.in <- pos.comp.out
map.scale.out <- structure(list(fill = c("#F8766D", "#00BA38", "#00BA38", "#619CFF"
), x = c(-1.16594083365053, -2.22008708454669, -2.83662163671106,
-3.9848822819069), y = c(5.13009568769485, 6.33635502737016,
6.8237884586677, 8.21105636898428), PANEL = structure(c(1L, 1L,
2L, 2L), .Label = c("1", "2"), class = "factor"), group = structure(c(1L,
2L, 2L, 3L), n = 3L)), class = "data.frame", row.names = c(NA,
-4L))
diffobj::diffPrint(map.scale.in, map.scale.out[names(map.scale.in)])
```

The fill aesthetic is transformed into color codes.  We reordered the output
columns in the diff to be in the same order as the input for clarity.

### Geom$use_defaults

#### Description

For each layer, add the geom default aesthetics and aesthetics specified as
non-mapping parameters to the [layer function](#layer-function) to the data.  It
is worth noting that any aesthetics added by this mechanism are not [mapped by
the scales](#map-scale).

#### Signature

* `$use_defaults([self], data, params)`

The `self` parameter, if present, can appear anywhere in the signature.  The
other parameters must be in the relative order shown above.  See [`ggproto`
methods](#ggproto-method).

#### Parameters

* `self`: optional parameter, if present will point to the `Geom` `ggproto`
  object this method belongs to.
* `data`: the layer data for.
* `params`: the parameter list as matched to the aesthetics supported by the
  `Geom` (i.e. `Geom$aesthetics()`).

#### Data Manipulation

Again, with the `geom_point` layer of our [example plot](#example-plot):

<!--
note we mess with the PANEL name to get around the bug with unable to ignore
whitespace around colon in diffStr
-->
```{r, echo=FALSE, results="asis"}
defaults.in <- map.scale.out
names(defaults.in)[names(defaults.in) == 'PANEL'] <- 'PANEL '
defaults.out <- structure(list(fill = c("#F8766D", "#00BA38", "#00BA38", "#619CFF"
), x = c(-1.16594083365053, -2.22008708454669, -2.83662163671106,
-3.9848822819069), y = c(5.13009568769485, 6.33635502737016,
6.8237884586677, 8.21105636898428), PANEL = structure(c(1L, 1L,
2L, 2L), .Label = c("1", "2"), class = "factor"), group = structure(c(1L,
2L, 2L, 3L), n = 3L), shape = c(21, 21, 21, 21), colour = c("black",
"black", "black", "black"), size = c(4, 4, 4, 4), alpha = c(NA,
NA, NA, NA), stroke = c(0.5, 0.5, 0.5, 0.5)), row.names = c(NA,
-4L), class = "data.frame")

diffobj::diffStr(defaults.in, defaults.out, mode='unified')
```

You can see how the data gains defaults (e.g `colour`, `alpha`, `stroke`) as
well as the `shape` and `size` [parameter aesthetics](#param-aesthetic) that
were specified in the `geom_point` call of the [example plot](#example-plot).

### Geom$draw_layer/panel/group {#geom-draw}

#### Description

These methods are responsible for [rendering](#render) the [layer
data](#layer-data) into [grobs](#grob).  Typically you would modify one of the
layer, panel, or group methods, listed here in order of increasing specificity.
If possible, you should modify the most specific method (e.g. the group method)
as that requires the least work.  If you need more control over the process you
can override one of the more general methods.  If you do this, for example by
overriding the panel method, you will be responsible for splitting the data into
groups, processing the groups, and re-assembling the result

`$draw_layer` is called on each layer of the plot.  The default
`$draw_layer` method will call `$draw_panel` on each [panel](#panel).
Finally the default `$draw_panel` method will call `$draw_group` on each
group in the [panel](#panel).

#### Signature

* `$draw_layer([self], data, params, layout, coord)`
* `$draw_panel([self], data, panel_params, coord, ...)`
* `$draw_group([self], data, panel_params, coord, ...)`

The `self` parameter, if present, can appear anywhere in the signature.  The
other parameters must be in the relative order shown above.  If you override
`$draw_panel`, or `$draw_group` methods you may add additional parameters after
the `coord` parameter.  These parameters will be automatically matched by and
passed on from the [layer function](#layer-function).

One of these methods should apply the [coordinate
transformation](#coordtransform) to the [position
aesthetics](#position-aesthetic).

#### Parameters

* `self`: optional parameter, if present will point to the `Stat` `ggproto`
  object the method belongs to.
* `data`: the [layer data](#layer-data) for `$draw_layer`, a subset of the
  layer data corresponding to a [panel](#panel) for `$draw_panel`, or a
  subset of the panel data corresponding to a group for `$draw_group`.
* `params`: the parameter list as matched to the `$draw_panel/group` methods
  by the [layer function](#layer-function).
* `panel_params`: the list computed in
  [`Layout$setup_panel_params`](#layoutsetup_panel_params).
* `coord`: the plot's [`Coord`](#ggproto-coord) [`ggproto` object](#ggproto).

#### Data Manipulation

This is the data that `$draw_panel` starts with on the first [panel](#panel) of
the `geom_point` layer of the [example plot](#example-plot):

```{r echo=FALSE}
defaults.out
```

It then returns a [`grob`](#grob):

```
List of 7
 $ x   : 'unit' num [1:2] 0.8309native 0.0455native
  ..- attr(*, "valid.unit")= int 4
  ..- attr(*, "unit")= chr "native"
 $ y   : 'unit' num [1:2] 0.356native 0.429native
  ..- attr(*, "valid.unit")= int 4
  ..- attr(*, "unit")= chr "native"
 $ pch : int [1:2] 21 21
 $ size: 'unit' num 1char
  ..- attr(*, "valid.unit")= int 18
  ..- attr(*, "unit")= chr "char"
 $ name: chr "geom_point.points.852"
 $ gp  :List of 4
  ..$ col     : chr [1:2] "#000000FF" "#000000FF"
  ..$ fill    : chr [1:2] "#F8766DFF" "#00BA38FF"
  ..$ fontsize: num [1:2] 12.3 12.3
  ..$ lwd     : num [1:2] 0.945 0.945
  ..- attr(*, "class")= chr "gpar"
 $ vp  : NULL
 - attr(*, "class")= chr [1:3] "points" "grob" "gDesc"
 ```

### Coord$transform

#### Description

Transforms non-Cartesian coordinates into Cartesian coordinates,
and then rescales them to fit in the 0-1 range as that is what [`grid`](#grid)
uses for its viewports.  This is akin to what happens to the non-[position
aesthetics](#position-aesthetic) in the [scale map](#train-map-scale) step.

This [method](#ggproto-method) should be applied by one of the
[`Geom$*_draw`](#geom-draw) methods.  Which one applies it will depend on the
`Geom`.

#### Data Manipulation

`GeomPoint` applies the coordinate transformation at the `$draw_panel` step, so
here is what it does to the first panel of the  `geom_point` layer in our
[example plot](#example-plot):

```{r, echo=FALSE, results="asis"}
coord.trans.in <- structure(list(fill = c("#F8766D", "#00BA38"), x = c(-1.16594083365053,
-2.22008708454669), y = c(5.13009568769485, 6.33635502737016),
    PANEL = structure(c(1L, 1L), .Label = c("1", "2"), class = "factor"),
    group = 1:2, shape = c(21, 21), colour = c("black", "black"
    ), size = c(4, 4), alpha = c(NA, NA), stroke = c(0.5, 0.5
    )), class = "data.frame", row.names = c(NA, -2L), vars = "PANEL")
coord.trans.out <- structure(list(fill = c("#F8766D", "#00BA38"), x = c(0.830902392317063, 
0.0454545454545454), y = c(0.35636943561787, 0.429476062264858
), PANEL = structure(c(1L, 1L), .Label = c("1", "2"), class = "factor"), 
    group = 1:2, shape = c(21, 21), colour = c("black", "black"
    ), size = c(4, 4), alpha = c(NA, NA), stroke = c(0.5, 0.5
    )), row.names = c(NA, -2L), vars = "PANEL", class = "data.frame")
diffobj::diffPrint(coord.trans.in, coord.trans.out)
```

## Definitions

These definitions are unofficial.  Some of them were made up by yours truly to
make it easier to reference recurrent concepts that lack formal definitions.

<dl>
<!---------------------------------------------------------------------------->
<dt><a name='aesthetic'></a>Aesthetic</dt>
<dd>
Strictly speaking in the context of `ggplot2` an aesthetic is a graphical
property of a plot, such as position, color, size, or line-type.  Most of the
time when we say "aesthetic" we mean "aesthetic variable".  These are variables
that contain data that is to be represented as an aesthetic of the plot.  For
example the `x` aesthetic variable will contain values that are
intended to represent a position along the `x` axis in a plot.

Aesthetic variables come about in three ways in `ggplot2`:

* <a name='mapped-aesthetic'></a>**Mapped aesthetics** are specified with
  `aes(...)` via the `mapping` parameter to `ggplot()` or [layer
  functions](#layer-function) (e.g. `ggplot(data) +
  geom_point(mapping=aes(x=time, y=value))` where the mapped aesthetics are `x`
  and `y`.
* <a name='stat-aesthetic'></a>**Stat aesthetics** are generated by the [stat
  computation](#stat-compute) process.  They end up as additional variables in
  the computed layer data, and can be referenced with the `stat` call in the
  mapping specification (e.g. `stat_bin(aes(y=stat(ncount)))`).  Which stat
  aesthetics are generated depends on the statistic.  Sometimes this is
  documented in the [layer function](#layer-function) that applies the stat
  (e.g.  `?stat_bin`), but not always.  If the variables are not documented, you
  can look at the most specific [`Stat$compute_*`](#stat-compute) method that
  the [`Stat`](#ggproto-stat) defines.
* <a name='param-aesthetic'></a>**Parameter aesthetics** are specified in [layer
  functions](#layer-function) as part of `...`.  For example, in
  `geom_point(mapping=aes(x=val), size=3)`, `size` is considered a parameter
  aesthetic because it is not part of the `mapping=aes(...)` call, whereas `x`
  is a [mapped aesthetic](#mapped-aesthetic).  Unlike [mapped
  aesthetics](#mapped-aesthetic), parameter aesthetics are neither
  [transformed](#scaletransform_df) nor [mapped](#map-scale).  Instead, they are
  added at the end of the `ggplot_build` process by
  [`Geom$use_defaults`](#geomuse_defaults).
</dd>
<!---------------------------------------------------------------------------->
<dt><a name='position-aesthetic'></a>Aesthetic, Position</dt>
<dd>
[Aesthetic](#aesthetic) variables that correspond to position values.  `x` and
`y` are the most common ones, but there are many others such as `xmin`, `xmax`,
`xend`, and the corresponding `y` versions of those.

Position and non-position aesthetics are treated slightly differently in the
[`ggplot_build`](#ggplot_build) process.  This is primarily because position
aesthetics are subject to [coordinate](#ggproto-coord) transformations, whereas
non-position aesthetics are not.
</dd>
<!---------------------------------------------------------------------------->
<dt><a name='ggproto-coord'></a>Coord, `ggproto`</dt>
<dd>
[`ggproto`](#ggproto) object responsible for:

* Coordinate transformations of the [position aesthetics](#position-aesthetic).
* [Rendering](#render) of the plot/panel [grob](#grob) meta-elements such as
  gridlines, panel-outline, panel-background, etc.
</dd>
<!---------------------------------------------------------------------------->
<dt><a name='facet'></a>Facet</dt>
<dd>
Facets are sub-viewports in which a subset of the data is displayed.  By default
plots have one facet that comprises the plot [panel](#panel), axes, and axis
labels.  More facets can be added by subdividing the data across
[panels](#panel) with the `facet_*` facet specification functions.

Each plot [layer](#layer) can potentially cover all plot facets, although some
layers may not have data to show in every facet.
</dd>
<!---------------------------------------------------------------------------->
<dt><a name='ggproto-facet'></a>Facet, `ggproto`</dt>
<dd>
[`ggproto`](#ggproto) object that:

* Computes [facet](#facet) layout.
* Assigns data to [panels](#panel).
* [Renders](#render) the facets by combining the panels with the layer
  [grobs](#grob), axes, labels, facet labels, etc., into a [`gtable`](#gtable).
</dd>
<!---------------------------------------------------------------------------->
<dt><a name='ggproto-geom'></a>Geom, `ggproto`</dt>
<dd>
`Geom*` [`ggproto`](#ggproto) objects such as `GeomPoint` or `GeomRect` are
responsible for translating the [layer data](#layer-data) into graphical
objects.  In `ggplot` these graphical objects are called
[grobs](#grob).  In order to create your own geoms you will
need to learn how to use `grid`, or alternatively to re-use existing `Geom*`
objects to generate grobs suited for your purposes.

See `vignette(package='ggplot2', 'extending-ggplot2')` for examples of how to
extend `Geom` objects.
</dd>
<!---------------------------------------------------------------------------->
<dt><a name='ggproto'></a>Ggproto</dt>
<dd>
Object oriented framework used by `ggplot2`.  See `?ggplot2::ggproto` for
details.  `ggproto` objects contain functions (a.k.a.
[methods](#ggproto-method)) and data, both of which can be accessed with the `$`
or `[[` operators.  Common `ggproto` objects include:

* [`Coord`](#ggproto-coord).
* [`Geom`](#ggproto-geom).
* [`Layer`](#ggproto-layer) (not exported).
* [`Layout`](#ggproto-layout).
* [`Position`](#ggproto-position).
* [`Scale`](#ggproto-scale).
* [`Stat`](#ggproto-stat).

`ggproto` objects are [reference objects](#reference), so they can be
modified in place by their [methods](#ggproto-method).

</dd>
<!---------------------------------------------------------------------------->
<dt><a name='ggproto-method'></a>Ggproto Method</dt>
<dd>
A function that is part of a [`ggproto`](#ggproto) object.  Typically these are
called with the `$` operator:

```{r eval=FALSE}
Object$method()
```

A feature of `ggproto` methods is that they can have a special `self`
parameter.  Methods that have the `self` parameter will be passed the
[`ggproto`](#ggproto) object they belong to as the value of that parameter.
This gives the method access to the object data and other methods, and the
opportunity to modify the [`ggproto` object](#ggproto) by reference as these are
[reference objects](#reference).

We illustrate the use of `self` with a simple example.  None of the method calls
provides the `self` parameter directly:

```{r eval=TRUE}
Object <- ggplot2::ggproto(
  "Example", NULL,
  method1=function(self, x) cat(sprintf("%s %s\n", x, self$secret)),
  method2=function(x) cat(sprintf("%s %s\n", x, self$secret)),
  method3=function(self, x) self$secret <- x,
  secret="secret"
)
Object$method1("big")  # works, `self` is provided automatically
Object$method2("big")  # Error because `self` not in signature
Object$method3("NEW SECRET")  # update `secret` member by reference
Object$method1("big")  # `secret` member updated
```

`ggproto` methods are generated at call time by the `$` operator for `ggproto`
objects.  The return value of `$` is a wrapper around the function provided to
the `ggplot2::ggproto` call.  It is this wrapper that passes the `self` object:

```{r}
Object$method1
```

This 'wrapping' behavior requires us to circumvent `$` and `[[` if we want
direct access to the method, e.g. for setting the `debug` state:

```{r, eval=FALSE}
debug(get("method1", Object))
Object$method1("big")  # this will be a debugged version of `method1`
```

This works because [`ggproto` objects](#ggproto) are environments.
</dd>
<!---------------------------------------------------------------------------->
<dt><a name='grid'></a>Grid</dt>
<dd>
Low level graphics package that comes packaged by default with R.  It is used
by `ggplot` to draw the plots.  There are several useful vignettes you can
find with `browseVignettes('grid')`.
</dd>
<!---------------------------------------------------------------------------->
<dt><a name='grob'></a>Grid Graphical Objects, a.k.a. Grobs</dt>
<dd>
R objects that contain a description of graphical elements that
can be used by the `grid` package to draw them on a graphical device.

See `?grid::grob`.
</dd>
<!---------------------------------------------------------------------------->
<dt><a name='gtable'></a>Gtable</dt>
<dd>
An object containing a collection of [grobs](#grob) intended to
be displayed in a table-like layout.  `gtable` objects are used by `ggplot2` to
simplify the manipulation of the layout of plots.  `gtable` objects are
themselves [`grobs`](#grob) and can be drawn by [`grid`](#grid).

See `?gtable::gtable`.
</dd>
<!---------------------------------------------------------------------------->
<dt><a name='layer'></a>Layer</dt>
<dd>
`ggplot2` plots are rendered in layers, one on top of each other.  This makes it
possible to have multiple super-imposed graphical representations of data.  For
example, we can overlay a representation of a model fit on top of a scatter-plot
of the data the model was trained on.

Each layer will be rendered across all [panels](#panel) it contains data for.

Internally, layers are represented by [`Layer` `ggproto`](#ggproto-layer)
objects.
</dd>
<!---------------------------------------------------------------------------->
<dt><a name='ggproto-layer'></a>Layer, `ggproto`</dt>
<dd>
[`ggproto`](#ggproto) object responsible for computing and rendering plot
[layers](#layer).  It contains [`Geom`](#ggproto-geom), [`Stat`](#ggproto-stat),
and [`Position`](#ggproto-position) `ggproto` sub-objects, along with other data
and the methods that carry out the computation and translation of layer data
into [grobs](#grob).

The `Layer` object is not exported and as such not part of the `ggplot2`
extension API.

See the [plot structure](#plot-structure) for more details.
</dd>
<!---------------------------------------------------------------------------->
<dt><a name='layer-data'></a>Layer Data</dt>
<dd>
The data associated with a particular `Layer`.  This is a `data.frame`.  Layers
that do not have a `data.frame` explicitly associated by the [layer
function](#layer-function) will typically inherit the [plot data](#plot-data).
</dd>
<!---------------------------------------------------------------------------->
<dt><a name='ggproto-layout'></a>Layout, `ggproto`</dt>
<dd>
[`ggproto`](#ggproto) object that contains all the metadata required to lay out
the plot across the [facets](#facet), and the methods to render the facet,
panel, and layer [grobs](#grob) together into a [gtable](#gtable).

Some useful methods/members:

* `Layout$coord`: the plot's [`Coord`](#ggproto-coord) [`ggproto`
  object](#ggproto).
* `Layout$facet`: the plot's [`Facet`](#ggproto-facet) [`ggproto`
  object](#ggproto).
* `Layout$get_scales`: a method which when given a panel id will return a list
  with the [position scales](#ggroto-scales-position) associated with that panel.
* [`Layout$layout` member](#layout-layout): a `data.frame` mapping panels to
  scales and positions in the overall plot.
* `Layout$render`: creates a [`gtable`](#gtable) by combining the layer, facet,
  and coordinate [grobs](#grob).
</dd>
<!---------------------------------------------------------------------------->
<dt><a name='layer-function'></a>Layer Function</dt>
<dd>
Functions like `geom_point` or `stat_bin` that are used to add layers to a
`ggplot`.  These functions return [`Layer` `ggproto`](#ggproto-layer) objects
that contain the layer specification, including the [`Geom`](#ggproto-geom),
[`Stat`](#ggproto-stat), and [`Position`](#ggproto-position) `ggproto`
sub-objects, and other relevant data.

A layer function's primary objectives are to:

* Translate shorthand geom/stat/position strings into the actual [ggproto
  objects](#ggproto).
* Assign its parameters to the [`Layer`](#ggproto-layer) object and sub-objects.

To illustrate consider the following call to a layer function:
```{r eval=FALSE}
ggplot2::stat_bin(geom='point', shape=21, bins=10)
```
The layer function converts the shorthand "point" into a `GeomPoint` `ggproto`
object, assigns the `shape` parameter to the [parameter
aesthetic](#param-aesthetic) list, and assigns the `bins` parameter to the
`StatBin` `ggproto` sub-object.

Parameter assignment is done by matching against the
`Geom`/`Stat` `$[compute|draw]_[panel|group]` method arguments.  For example,
`StatBin` has both `$compute_panel` and `$compute_group` methods.  The layer
function will look at the parameters of the `$compute_panel` method, and if that
method has a `...` parameter as is the case here, it will also look at the
parameters of the `$compute_group` method.  The primary implication of this
behavior for extension writers is that when you override
`$[compute|draw]_[panel|group]` in a [`Geom`](#ggproto-geom) or
[`Stat`](#ggproto-stat), any parameters your method has will
automatically be matched by the layer function that uses it.

Layer function parameters are also matched against that layer's
[`Geom`](#ggproto-geom) aesthetics.  You can see what those are using the
`$aesthetics` method:

```{r}}
ggplot2::GeomPoint$aesthetics()
```

Layer function parameters that match more than one of the
[`Stat`](#ggproto-stat) methods, [`Geom`](#ggproto-geom) methods, or
[`Geom`](#ggproto-geom) [aesthetics](#aesthetic) are copied for each of the
layer elements they match.
</dd>
<!---------------------------------------------------------------------------->
<dt><a name='panel'></a>Panel</dt>
<dd>
Each plot layer is displayed across the plot [facets](#facet).  The part
of the facet that displays the layer data is the panel.  If a facet is a window,
then the panel is the glass pane.

Internally, each panel corresponds to a subset of the rows of the layer data.
The `PANEL` variable added to the [layer data](#layer-data) by
[`Layer$layer_data`](#layerlayer_data) designates what panel each row in the
layer data should be displayed in.
</dd>
<!---------------------------------------------------------------------------->
<dt><a name='plot-data'></a>Plot Data</dt>
<dd>
The `data.frame` specified in the top level `ggplot(data=...)` call is the plot
data.  [Layers](#layer) that are not explicitly assigned [layer
data](#layer-data) inherit this `data.frame` as their data.
</dd>
<dt><a name='plot-spec'></a>Plot Specification</dt>
<dd>
The result of a `ggplot(...) + ...` call.  This is a `ggplot` S3 object that
is initialized with the plot data and any additional instructions the user
provided, such as layers and [aesthetic mappings](#mapped-aesthetic),
[scales](#ggproto-scale), themes, etc.

See the [plot structure](#plot-structure) section for details.
</dd>
<!---------------------------------------------------------------------------->
<dt><a name='ggproto-position'></a>Position, `ggproto`</dt>
<dd>
`Position*` [`ggproto`](#ggproto) objects such as `PositionDodge` are
responsible for adjusting the position aesthetics in the layer data.  For
example, `PositionDodge` shifts and adjusts `x` [position
aesthetics](#position-aesthetic) such that objects overlapping on the `xmin`
aesthetic are placed side by side and reduced in width to fit where one of them
would have it previously.

See [`Position$compute_*`](#position-compute) for details and `?Position`.
</dd>
<!---------------------------------------------------------------------------->
<dt><a name='reference'></a>Reference Object</dt>
<dd>
Objects that are tied to a particular memory reference, and as such break the
default pass-by-value semantics of R.  This means that an object provided to
a function and modified in that function's scope will appear as modified in the
calling scope (and everywhere else it is referenced).  Environment objects are
of this type, whereas list are not:

```{r}
env <- new.env()
lst <- list()
env$a <- lst$a <- 1

f <- function(e, l) {
  e$a <- 2
  l$a <- 2
}
f(env, lst)

env$a
lst$a
```
[`ggproto` objects](#ggproto) are reference objects.  They are implemented as
environments with a class attribute.
</dd>
<!---------------------------------------------------------------------------->
<dt><a name='render'></a>Render</dt>
<dd>
In `ggplot2`, the terms "draw" and "render" are used to denote the process of
creating [grobs](#grob) or [`gtable` objects](#gtable). This is different to the
more traditional definition of converting the graphical object models (e.g. the
[grobs](#grob)/[gtable](#gtable)) into a 2D image.
</dd>
<!---------------------------------------------------------------------------->
<dt><a name='ggproto-scale'></a>Scale</dt>
<dd>
[`ggproto` object](#ggproto) that Contains meta data about the scales (e.g. the
range of values plotted under the scale) and [methods](#ggproto-method) to
transform the data.

For example, `scale_x_sqrt` produces a `ScaleContinuous` [`ggproto`](#ggproto)
object with a [`$transform` method](#scaletransform_df) that will covert any `x`
related aesthetics to their square roots.

There are is also typically a [`Scale$map`](#train-map-scale) method to [map the
aesthetics](#map-scale) into the display space (e.g. categories into colors),
although for [position aesthetics](#position-aesthetic) that is typically
handled by [`Coord$transform`](#coordtransform).
</dd>
<!---------------------------------------------------------------------------->
<dt><a name='ggproto-scale-position'></a>Scale, Position</dt>
<dd>
[Scales](#ggproto-scale) that correspond to the [position
aesthetics](#position-aesthetic).
</dd>
<!---------------------------------------------------------------------------->
<dt><a name='map-scale'></a>Scale, Map</dt>
<dd>
Use [scales](#ggproto-scale) to convert aesthetic values into the domain they
will be plotted in.  For example, values on a color/fill scale are converted to
the colors they will be displayed as.  Mapping for [position
scales](#ggproto-scale-position) would in theory convert values into the 0-1
range, but that is actually handled by the [coordinate
transformation](#coordtransform) instead.
</dd>
<!---------------------------------------------------------------------------->
<dt><a name='train-scale'></a>Scale, Train</dt>
<dd>
Process that computes the range of values covered by a scale.  For continuous
scales this is typically just the min and max values, whereas for discrete
scales it is the sorted unique listing of all the different values.

Typically [position scales](#ggproto-scale-position) are trained several times
in the plot build process as they are used therein and the data is changed, e.g.
via statistics or position adjustments.
</dd>
<!---------------------------------------------------------------------------->
<dt><a name='ggproto-stat'></a>Stat</dt>
<dd>
`Stat*` [`ggproto`](#ggproto) objects such as `StatBin` are responsible for
[recomputing](#stat-compute) the layer data with a statistic and potentially
adding new variables.  For example, `StatBin` bins the data and computes `y`
values as well as other stats such as `count` and `ncount`.

See `vignette(package='ggplot2', 'extending-ggplot2')` for examples.
</dd>
</dl>
